---
title: "Unsupervised Learning with Commodity Prices"
date: Last updated on `r Sys.Date()`
output: 
  github_document:
    toc: TRUE
    toc_depth: 4
---

## Summary

This project attempts to utilize methods in Unsupervised Learning to deduce and infer similarities in the trading behaviour of commodity contracts. Specifically, it uses Hierarchical Clustering and K-Means Algorithms to investigate if different commodity contracts may be clustered into reasonable sub-groups.

## Package Installs

```{r install_packages, message = FALSE, warning = FALSE}
library(cluster)
library(dendextend)
library(naniar)
library(Quandl)
library(tidyverse)
library(xts)
```

## Code

### 1. Import Data

```{r import_quandl}
# Load the Quandl API Key
Quandl.api_key("rn2xyN_hG9XfxN_9ibFJ")

# Define a vector of Commodity Names
names <- c("Brent_Crude", "WTI_Crude", "Dubai_Crude", "Canadian_Crude", "HHB_Gas", "NBP_Gas", "RBOB_Gasoline", "European_Naphtha", "GulfCoast_Jet", "Singapore_FO", "CME_Copper", "SHFE_Copper", "TSI_Iron_Ore", "SGX_Iron_Ore", "Newcastle_Coal", "Indonesian_Coal", "CME_Aluminium", "SHFE_Aluminium")

# Define a vector of Quandl Codes per Commodity
codes <- c("CHRIS/ICE_B1", "CHRIS/CME_CL1", "CHRIS/CME_DC1", "CHRIS/CME_WCC1", "CHRIS/CME_NG1", "CHRIS/ICE_M1","CHRIS/CME_RB1", "CHRIS/CME_UN1", "CHRIS/CME_GE1", "CHRIS/CME_UA1", "CHRIS/CME_HG1", "CHRIS/SHFE_CU1", "CHRIS/CME_TIO1", "CHRIS/SGX_FEF1", "CHRIS/ICE_NCF1", "CHRIS/CME_MCC1", "CHRIS/CME_ALI1", "CHRIS/SHFE_AL1")

# Create a data.frame to store commodity contract codes
commod_info <- data.frame(
  name = names,
  code = codes,
  stringsAsFactors = FALSE
)

# View the first 6 rows of commodities
head(commod_info)
```
```{r import_data}
# Create an empty list to store historical prices per commodity: df_list
df_list <- list()

# Iterate through commod_info and add to df_list
for (i in 1:nrow(commod_info)) {
  df_list[[i]] <- Quandl(commod_info$code[i])
}
```

### 2. Preprocess Data

```{r merge_datasets}
# Reduce all data.frames in df_list into a single object: com
com <- df_list %>% 
  reduce(left_join, by = "Date") %>% 
  select(matches("Date|Settle")) %>% 
  select(-"Pre Settle.x", -"Pre Settle.y")

# Define column names as full commodity name
names(com) <- c("Date", commod_info$name)

# Impute missing NA values with LOCF
com <- na.locf(com)

# View the first 10 observations of com
head(com, n = 10)
```
```{r observe_str}
# Observe the structure of the joined data.frame
str(com)
```

From this early pre-processing step, we have a data.frame (com) consisting of a Date Index, as well as 18 variables/columns for each commodity under analysis. To generate the distance matrix for Hierarchical Clustering, we will first need to reduce the data.frame and transpose it such that the time index is reflected in the columns. From a tidyverse perspective, we can then interpret each row (each Commodity Index) as one observation, and each column (each Date Index) as an attribute/variable corresponding to that observation.

We utilize the xts package in this case for easy pre-processing of the data.frame by time.

```{r scale_prices}
# Scale all prices in the data.frame: scaled_com
scaled_com <- com %>%
  select(-Date) %>%
  scale() %>%
  as.data.frame() %>%
  mutate(Date = com$Date) %>%
  select(Date, everything())

# Observe the first 6 observations of scaled_com
head(scaled_com)
```
```{r convert_to_xts}
# Convert the data.frame to an xts object: com_xts
com_xts <- as.xts(scaled_com[, -1], order.by = as.Date(com$Date))

# Subset the xts object for data after 2016: com_xts_sub
com_xts_sub <- com_xts['2016-01/2019-01']

# Convert the daily series to a monthly series without OHLC: com_xts_monthly
com_xts_monthly <- to_period(com_xts_sub, period = "months", OHLC = FALSE)

# View the first 6 observations of the xts object
head(com_xts_monthly)
```
```{r viz_prices, fig.height=5, fig.width=8, fig.align="center"}
# Convert the xts object back to a data.frame: com_monthly_df
com_monthly_df <- as.data.frame(com_xts_monthly)

# Visualize the price action of all commodities
com_monthly_df %>%
  mutate(Date = as.Date(rownames(com_monthly_df))) %>%
  select(Date, everything()) %>%
  gather(Commodity, Price, -Date) %>%
  ggplot(aes(Date, Price, color = Commodity)) +
    geom_line() +
    theme_minimal() +
    labs(
      title = "Historical Prices of Commodity Contracts",
      subtitle = "Scaled price action of 18 commodities from 2016 to 2019",
      caption = "Source: Quandl Continuous Futures Wiki",
      x = "Date",
      y = "Scaled Price"
    )
```

From this visualization from prices, it is not immediately obvious which set of commodities move in tandem with one another. We can, however, identify some salient facts such as the outlier status of the CME Aluminium Contract.

As the final stage of pre-processing, we transpose the data.frame to show time as columns and commodities as rows.

```{r transpose_df}
# Transpose the data.frame s.t. Dates are columns: com_final
com_t <- as.data.frame(t(com_monthly_df))

head(com_t)
```

### 3. Hierarchical Clustering

```{r generate_dendrogram, fig.height=5, fig.width=8, fig.align="center"}

# Define a theoretical number of clusters: k
K = 4

# Calculate a Distance Matrix between each Commodity: dist_com
dist_com <- dist(com_t, method = "euclidean")

# Generate an average linkage analysis: hc_com
hc_com <- hclust(dist_com, method = "average")

# Create a dendrogram from hc_com: dend_com
dend_com <- as.dendrogram(hc_com)

# Plot the dendrogram
dend_colored <- color_branches(dend_com, k = K)
plot(dend_colored, main = "Clustered Dendrogram of Commodity Price Action")
```
```{r tidy_with_hc}
# Convert commodity IDs from rownames to a new column: Commodity
df_com <- rownames_to_column(as.data.frame(com_t), var = "Commodity")

# Create cluster assignments with h = 5
cut_com <- cutree(hc_com, k = K)

# Create a new column reflecting cluster assignments from cut_com
clustered_com_hc <- mutate(df_com, Cluster = cut_com)

# Create a tidy data.frame by gathering the Month and Values into two columns
tidy_com_hc <- gather(
  data = clustered_com_hc,
  key = Month,
  value = Price,
  -Commodity,
  -Cluster
)

# View the first 6 observations of tidy_com
head(tidy_com_hc)
```
```{r view_clusters_hc}
# View cluster assignments
sort(cut_com)
```

```{r viz_hc_results, fig.height=5, fig.width=8, fig.align="center"}
# Visualize each time series by Cluster Assignment
tidy_com_hc %>% 
  mutate(Month = as.Date(Month)) %>% 
  ggplot(aes(Month, Price, color = factor(Cluster))) +
    geom_line(aes(group = Commodity)) +
    scale_color_manual(values = c(
      "#999999", 
      "#E69F00", 
      "#56B4E9", 
      "#009E73", 
      "#F0E442")) +
    labs(
      title = "Hierarchical Clusters of Traded Commodity Contracts",
      subtitle = "Based on Historical Price Action from 2016 to 2019",
      caption = "Source: Quandl Continuous Futures Wiki",
      x = "Date",
      y = "Scaled Price",
      color = "Cluster"
    ) +
    theme_classic()
```

### 4. K-Means Clustering: K Estimation

```{r calculate_totwithinss}
# Estimate Total Within Sum of Squares per value of k
tot_withinss <- map_dbl(1:10, function(k) {
  model <- kmeans(x = com_t, centers = k)
  model$tot.withinss
})

tot_withinss
```
```{r viz_elbow}
# Create a data.frame containing both k and tot_withinss: elbow_df
elbow_df <- data.frame(
  k = 1:10,
  tot_withinss = tot_withinss
)

# Visualize the Elbow Plot with ggplot
elbow_df %>% 
  ggplot(aes(k, tot_withinss)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Elbow Plot of Estimated Commodity Clusters",
    subtitle = "Total Within Sum of Squares per value k",
    caption = "Source: Quandl Continuous Futures Wiki",
    x = "Num of Clusters",
    y = "Total Within Sum of Squares"
  ) +
  theme_minimal()
```

Analysis of the Elbow Plot suggests that both k = 2 and k = 4 are good contenders for estimated number of clusters.

```{r calculate_sil}
# Estimate Silhouette Width per unit k
sil_width <- map_dbl(2:10, function(k) {
  model <- pam(com_t, k = k)
  model$silinfo$avg.width
})

sil_width
```
```{r viz_sil}
# Create a data.frame containing both average silhouette wdith and k: sil_df
sil_df <- data.frame(
  k = 2:10,
  sil_width = sil_width
)

# Visualize the relationship between k and silhouette width with ggplot
sil_df %>% 
  ggplot(aes(k, sil_width)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 2:10) +
  labs(
    title = "Silhouette Width Plot of Estimated Commodity Clusters",
    subtitle = "Average Silhouette Width Per Number of Clusters k",
    caption = "Source: Quandl Continuous Futures Wiki",
    x = "Number of Clusters",
    y = "Average Silhouette Wdith"
  ) +
  theme_minimal()
```

Similar to the Elbow Plot, analysis of the Average Silhouette Width per number of clusters suggests that k = 2, k = 4 and k =5 are all good estimations of the number of clusters.

Given the similarities in the Elbow Plot and Average Silhouette Width plot, we theorize k = 4 for our K-Means Clustering analysis.

### 5. K-Means Clustering: Cluster Assignment

```{r km_clustering}
# Build a K-Means Model with k = 4
model_com_km <- kmeans(com_t, centers = 4)

# Extract the vector of cluster assignments from the model
clust_kmeans <- model_com_km$cluster

# View the results of cluster assignment with K-Means
clust_kmeans
```
```{r tidy_with_km}
# Create a new column reflecting cluster assignments from clust_kmeans: clustered_com_km
clustered_com_km <- mutate(df_com, Cluster = clust_kmeans)

# Create a tidy data.frame by gathering the Month and Values into two columns
tidy_com_km <- gather(
  data = clustered_com_km,
  key = Month,
  value = Price,
  -Commodity,
  -Cluster
)

# View the first 6 observations of tidy_com_km
head(tidy_com_km)
```
```{r view_km_results}
# View cluster assignments with K-Means Clustering
sort(clust_kmeans)
```
```{r viz_kmeans}
# Visualize each time series by Cluster Assignment
tidy_com_km %>% 
  mutate(Month = as.Date(Month)) %>% 
  ggplot(aes(Month, Price, color = factor(Cluster))) +
    geom_line(aes(group = Commodity)) +
    scale_color_manual(values = c(
      "#999999", 
      "#E69F00", 
      "#56B4E9", 
      "#009E73", 
      "#F0E442")) +
    labs(
      title = "K-Means Clusters of Traded Commodity Contracts",
      subtitle = "Based on Historical Price Action from 2016 to 2019",
      caption = "Source: Quandl Continuous Futures Wiki",
      x = "Date",
      y = "Scaled Price",
      color = "Cluster"
    ) +
    theme_classic()
```

### 6. Conclusions

This preliminary exercise with Hierarchical and K-Means Clustering appears to suggest that: of the 18 selected commodity contracts, *there are broadly 4 clusters of commodities* whose price action resemble one another the most. 

#### Cluster 4: CME Alumnium
One example of such a cluster is the Chicage Mercantile Exchange Aluminium Contract, whose price behaviour is distinctly different from that of the other 17 commodities under analysis. In both the Hierarchical Clustering and K-Means Clustering Analysis, its price action stands out from the rest.

#### Cluster 2: Brent, WTI, Henry Hub Gas, NBP Gas
With respect to the analysis of individual clusters, an interesting hypothesis would be that Cluster 2 Commodities (E.g. Brent, WTI, Henry Hub, NBP, Gasoline, Copper etc.) are the commodities that depend more heavily on Macro factors. Historically, these assets (as benchmark commodity prices) tend to move more in tandem with the state of the global macroeconomic environment relative to assets found in other clusters (E.g. European Naphtha and Chinese Iron Ore).

#### Cluster 3: Dubai Crude, SGX Fuel Oil, SHFE Copper, Iron Ore
In the case of Cluster 3 commodities such SHFE Copper, SGX Fuel Oil and Iron Ore, their relationship to the Macro environment is less obvious. In fact, it could be argued that these commodities are more dependent on individual supply-demand dynamics as well as regional factors (E.g. IMO2020 for Fuel Oil prices and Chinese Government Policy for Iron Ore).

#### Use-Cases
Potential use-cases of this analysis would be to potentially design pairs-trading strategies around intra-cluster commodities as well as to use the movement of one asset as a predictor of another.
